Scaffolder - Software for Reproducible Genome Scaffolding.

### 2011-03-02 src/article.tex
Proof read results

The Scaffolder software allows computational joining of nucleotide sequences
together into a single contiguous scaffolded super-sequence. Plain-text
scaffold files written in the YAML syntax specify how these sequences should be
joined and the scaffold file as read by the scaffolder software is used to
generate the scaffold sequence. In addition to specifying which contigs are
required, the scaffold file allows the contigs to be edited to produce smaller
sub-sequences or reverse complemented if necessary. The process of genome
finishing may involve producing additional oligonucleotide sequences to fill
unknown regions in the scaffold. The scaffold file format allows these
additional insert sequences to bridge these gaps. These inserts can also be
treated in a similar way to larger contig sequences: trimmed and/or reverse
complemented to match the corresponding gap region. \pb

The distances between contigs may be estimated from paired read data or from
mapping the contigs to a reference genome. These inter-contig gap regions are
useful to join separate sequences together by an estimated size and the
scaffold file allows for the specification of these unknown regions. The use of
these regions in the scaffold which areas in the scaffold and their approximate
size are still to be resolved. \pb

The scaffold file specifies how sequence regions are joined together; however
the sequences themselves are not stored in the scaffold file and are instead
maintained as a separate FASTA file. Series of nucleotide sequences in
a FASTA-formatted file is standard output for a genome assembler. These may be
contigs, sets of contigs scaffolded into larger sequences, or both. These
nucleotide sequences stored in the corresponding FASTA file are referenced in
the scaffold using the first word form the FASTA header of the corresponding
nucleotide sequence. Maintaining the nucleotide sequences in a separate file
preserves the original sequence and decouples the original sequence data from
the scaffold file describing how it should be used to in the genome sequence.
\pb

@@

The scaffold file is written using the YAML syntax. An example scaffold file is
shown in Figure 1. This file illustrates the text attributes used to describe
a scaffold and the joining of sequences in the genome build. The basic layout
of the scaffold file is a list of entries, where each entry corresponds to
a region of sequence in the generated scaffold super-sequence. \pb

@@

The first line of the scaffold file begins with three dashes to indicate the
start of a YAML-formatted document. The first entry (highlighted in green)
begins with a dash character `-' to denote an entry in the YAML list. The first
line of each entry in the scaffold therefore begins with a dash line. The next
line is indented by two spaces as whitespace is required to group similar
attributes together. The ``sequence'' tag indicates that this entry corresponds
to a sequence in the FASTA file and the following line indicates the name of
this FASTA sequence using the ``source'' tag. As described the first word of
the FASTA header is used to identify which sequence is selected from the file.
Together these three lines describe the first entry in the scaffold as a simple
sequence using the FASTA sequence named `sequence1'. On the right hand side of
Figure 1 this produces the first region in the scaffold, also shown in green.
\pb

@@

The second entry in the scaffold, highlighted in orange, is identified by the
``unresolved'' tag, indicating a region of unknown sequence but known length.
The second line specifies the size of this unknown region. This entry therefore
inserts a region of 20 `N' characters in the scaffold. \pb

@@

The last entry in the scaffold, highlighted in blue, adds the FASTA entry
`sequence2' to the scaffold. This entry demonstrates how this sequence may be
manipulated prior to addition to the scaffold. The `start' and `stop' tags trim
the sequence to these coordinates inclusively, while the ``reverse'' tag
instructs scaffolder to also reverse complement the sequence. In the scaffolded
genome shown in Figure 1 this completes the scaffold sequence. \pb

The final entry in the scaffold also includes the ``inserts'' tag to add
additional regions of sequence. These inserts are also added as a YAML list,
with each insert beginning with a dash line. The first insert, shown in purple,
similar attributes to that of a sequence entry; the reverse, start and stop
tags are used to trim and reverse complement the insert. Similarly the `source'
tag identifies the corresponding FASTA sequence as `insert'. The ``open'' and
``close'' tags are specific to inserts and determine where the insert is added
in the enclosing sequence. The region of the sequence inside these coordinates
is inclusively replaced by the specified insert FASTA sequence. This is
visualised in the putative gene Figure 1 by the black lines bisecting the blue
sequence. \pb

The next insert, shown in brown, is specified using only the `open' tag. This
illustrates that only one of either `open' or `close' tags is required when
adding an insert sequence. If only only one of the `open' or `close' tags is
used the corresponding opposing position distance is calculated from the length
of the insert FASTA sequence. This allows inserts to bridge into, and partially
fill gap regions without requiring an end coordinate position. \pb

\subsection{Scaffolder Software Interface} %{{{2

Scaffolder provides a standardised set of Ruby classes and methods (API) for
interacting with the scaffold. This allows Scaffolder to be integrated into
existing genomics workflows or used with Ruby build tools such as Rake. In
addition Scaffolder provides a command line interface (CLI) providing functions
to validate the scaffold file and build a draft super sequence. The Scaffolder
CLI behaves as a standard Unix tool and returns appropriate exit codes and
manual pages. The use of both these Scaffolder interfaces is outlined in detail
on the scaffolder website (\scaffolder). \pb

@@

Scaffolder was designed to be as simple to use as possible and assuming the
Ruby and RubyGems software are present, Scaffolder can be installed in a single
command line step. This should negate any possible barriers to entry which
require manual compilation of source code. Scaffolder also uses a minimal and
compact syntax to describe the genome scaffold. This syntax is simple to
construct and edit whilst being succinct and readable. The goal of this syntax
is to make genome scaffolds easy to construct in a common text editor. \pb

Constructing a genome using a text file makes generating a scaffold super
sequence both reproducible and deterministic for the same scaffold and FASTA
sequence file. In comparison, joining large nucleotide sequences by hand cannot
be reliably reproduced. The scaffold file furthermore provides a human readable
description of how the scaffold is constructed. \pb

Configuring the final sequence in the scaffold file means the build is easier
to edit once already constructed. The YAML text format also allows comparison
of differences between scaffold builds using standard Unix tools such as diff.
This makes scaffold files easier to store in version control systems and allows
genome finishers methods similar to those in software development. \pb

>>>>

The Scaffolder software allows computational joining of nucleotide sequences
together into a single contiguous scaffolded super-sequence. Plain-text
scaffold files written in the YAML syntax specify how these sequences should be
joined and the scaffolder software is used to generate the scaffold sequence.
In addition to specifying which contigs are required, the scaffold file allows
the contigs to be edited into smaller sub-sequences or reverse complemented if
necessary. The process of genome finishing may involve producing additional
oligonucleotide sequences to fill unknown regions in the scaffold. The scaffold
file format allows these additional insert sequences to bridge gaps. These
inserts can also be treated in a similar way to larger contig sequences:
trimmed and/or reverse complemented to match the corresponding gap region. \pb

The distances between contigs may be estimated from paired-read data or from
mapping the contigs to a reference genome. These inter-contig gap regions are
useful to join separate sequences together by an estimated size and the
scaffold file allows for the specification of such unresolved regions. The use
of these regions in the scaffold indicates the unresolved regions in the build
and their approximate size. \pb

The scaffold file specifies how sequence regions are joined together; however
the sequences themselves are not stored in the scaffold file and are instead
maintained as a separate FASTA file. Nucleotide sequences in a FASTA-formatted
file are the standard output for a genome assembler, and these may be contigs,
sets of contigs scaffolded into larger sequences, or both. These nucleotide
sequences stored in the FASTA file are referenced in the scaffold using the
first word from in the FASTA header of the corresponding sequence. Maintaining
the nucleotide sequences in a separate file preserves the unedited sequence and
decouples the data from the scaffold file describing how it should be used to
in the genome sequence. \pb

@@

The scaffold file is written using the YAML syntax and an example is shown in
Figure 1. This file illustrates the text attributes used to describe a scaffold
and how the sequences are joined together in the genome build. The basic layout
of the scaffold file is a list of entries, where each entry corresponds to
a region of sequence in the generated scaffold super-sequence. \pb

@@

The first line of the scaffold file begins with three dashes to indicate the
start of a YAML-formatted document. The first entry (highlighted in green)
begins with a dash character `-' to denote an entry in the YAML list. This is
a requirement of the YAML format: that each entry begins with a dash line. The
next line is indented by two spaces as whitespace is required to group similar
attributes together. The ``sequence'' tag indicates that this entry corresponds
to a sequence in the FASTA file and the following line indicates the name of
this sequence using the ``source'' tag. As described the first word of the
FASTA header is used to identify which sequence is selected from the file.
Together these three lines describe the first entry in the scaffold as a simple
sequence using the sequence named `sequence1'. On the right hand side of Figure
1 this produces the first region in the scaffold, also shown in green. \pb

@@

The second entry in the scaffold, highlighted in orange, is identified by the
``unresolved'' tag, indicating a region of unknown sequence but known length.
The second line specifies the size of this unknown region. This entry produces
a region of 20 `N' characters in the scaffold. \pb

@@

The last entry in the scaffold, highlighted in blue, adds the sequence named
`sequence2' to the scaffold. This entry demonstrates how this sequence may be
manipulated prior to addition to the scaffold. The `start' and `stop' tags trim
the sequence to these coordinates inclusively, while the ``reverse'' tag
instructs scaffolder to also reverse complement the sequence. In the scaffolded
genome shown in Figure 1 this completes the scaffold sequence. \pb

The final entry in the scaffold also includes the ``inserts'' tag to add
additional regions of sequence. These inserts are also added as a YAML list,
with each insert beginning with a dash. The first insert, shown in purple,
similar attributes to that of a sequence entry; the reverse, start and stop
tags are used to trim and reverse complement the insert. Similarly the `source'
tag identifies the corresponding FASTA sequence as `insert1'. The ``open'' and
``close'' tags are specific to inserts and determine where the insert is added
in the enclosing sequence. The region of the sequence inside these coordinates
is inclusively replaced by the specified insert sequence. This is visualised in
the putative gene Figure 1 by the black lines bisecting the blue sequence. \pb

The next insert, shown in brown, is specified using only the `open' tag. This
illustrates that only one of either `open' or `close' tags is required when
adding an insert sequence. If only one of the `open' or `close' tags is used
the corresponding opposing position distance is calculated from the length of
the insert FASTA sequence. This allows inserts to bridge into, and partially
fill gap regions without requiring an end coordinate position. \pb

\subsection*{Scaffolder Software Interface} %{{{2

Scaffolder provides a standardised set of Ruby classes and methods (API) for
interacting with the scaffold. This allows Scaffolder to be integrated into
existing genomics workflows or used with Ruby build tools such as Rake. In
addition Scaffolder provides a command line interface (CLI) to validate the
scaffold file and build a draft super sequence. The Scaffolder CLI behaves as
a standard Unix tool and returns appropriate exit codes and manual pages. The
use of both these Scaffolder interfaces is outlined in detail on the scaffolder
website (\scaffolder). \pb

@@

Scaffolder was designed to be as simple to use and, assuming the Ruby and
RubyGems software are present, Scaffolder can be installed in a single command
line step. This should negate any possible barriers to entry which require
manual compilation of source code. Scaffolder also uses a minimal and compact
syntax to describe the genome scaffold. This syntax is simple to construct and
edit whilst being succinct and readable. The goal of this syntax is to make
genome scaffolds easy to make in a common text editor. The use of word
processors for this task is however discouraged. \pb

Constructing a genome using a text file makes generating a scaffold super
sequence both reproducible and deterministic for the same scaffold and FASTA
sequence file. In comparison, joining large nucleotide sequences by hand cannot
be reliably reproduced. The scaffold file furthermore provides a human readable
description of how the scaffold is constructed. Configuring the final sequence
in the scaffold file means the build is easier to edit once already
constructed. The YAML text format allows comparison of differences between
scaffold builds using standard Unix tools such as diff. This makes scaffold
files easier to store in version control systems and allows genome finishers to
use methods similar to those in software development. \pb
